Very Large Integers and polynomial inner products

### 2013-01-08 doc/paper_vli/vli.tex
some changes and TODO comments

%\date{}                                           % Activate to display a given date or no dater
\begin{document}
\maketitle
%\section{}
%\subsection{}
%\vspace{-0.65cm}
\begin{abstract}
The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 512 bits. 
In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library in assembly language for the CPU and the GPU.  All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach. We save a factor up to 10 on CPU, and 60 on GPU compared to a GMP solution,
and attain 67 \% of the peak performance of the GPU.
\end{abstract}
\section{Introduction}
 A few libraries like  LINBOX or NTL  provide high performance for polynomial solvers.
These libraries utilize with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
It is therefore necessary  to adapt the libraries to these new hardwares.  One example on this transition is the well know  blas/lapack libraries to MAGMA. 
Polynomials libraries, however  have not yet  do the transition to GPU.
During the last few years, the polynomial multiplication was successfully implemented on this new hardware
\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials,
 using a FFT approach.

In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
We preferred to develop a long integer library  rather than utilizing GMP, because GMP is a dynamic allocated memory library.
Moreover, we want a library natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated by meta-programming during the compilation for CPU and GPU.
We do not waste time in the dynamic memory allocation, and let the compiler optimize  the library for both architectures.

@@

The library performs the inner products of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
\begin{eqnarray}
P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
\end{eqnarray}
where  $x,\,y,\,z$ and $w$ are the variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner products  consist of the product 
of all entries  associated with a global reduction. 
 \begin{eqnarray}
 P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
\end{eqnarray}
where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
A final coefficient may be the sum of several coefficients (called contributions), resulting from the sum of several multiplications. These coefficients are the cross terms. 
The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.

@@

The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits; the size is a  template parameter, and
varies from 128 to 512 bits.  Its implementation is static, contrary to GMP. 
Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.

The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift. The library allows  operations between operands of the same type, and with basic  integers. For performance reasons, we have a specific solver for every operation except  for division. All the algorithms and operations are based on basic arithmetic  but the implementations have been done in assembly language to  boost  performance.

@@

We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a "carry bit" and an extended multiplication. The hardware supports
the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.

In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{sbbq/subq} for the subtraction, and \texttt{mul} for the extended multiplication.   
Combining these mnemonics, we reproduce easily long arithmetic algorithm, as described into the famous  ASM book of \cite{Hyde:2003:AAL:861534}.

We coded the kernels using the GNU inline assembly which consists of strings between ASM tags.
Generating by hand all these  assembly strings (kernels)  can be painful and a source of error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX} 
We designed a assembly generator. It creates all versions of a kernel during the compilation.   
It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.

@@

The x86-64 architecture allows 16 registers (32 for Power architecture). As much as possible, in the solver, we conserve the data in the registers in order to reduce memory read/writes to the  minimum.  Additional optimizations have been done like full unrolling and assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly language.
 PTX does not support carry bit for 64-bit integer, but only for 32-bit integer\footnote{Introduced since CUDA 4.2}.  
It allows multiplication-addition of 32-bit integer  with carry bit support \cite{CUDAasm}. 
Calculation in 32-bit mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.

@@

There exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA). Both have  advantages and disadvantages. In our problem the AoS consists in saving the data in this order : VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector). The SoA structure interleaves the coefficients of the VLI and the polynomials.

@@

The choice of the memory layout is important. It may facilitate  the usage of the SIMD/AVX instructions. We choose the AoS for the following reasons:
we want to reuse the VLI library without polynomial in other code, so we can not interleave the VLI number with the polynomial. 
Moreover  applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the multimedia register.
The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version. Moreover, there is a large overhead  to prepare the registers and shuffle the data into the registers. For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will not be the optimal solution. However SIMD has been used with success for  integers larger  than 1024-bit \cite{SIMD}.

@@

A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations Ð the same program 
is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) by group of 32 (warp).
The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
 and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block. 
% During the execution, independent blocks will be executed on free SMs, where the program of the block is executed over the threads concurrently way.

@@

Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALU  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).  We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time. For more realistic comparison the execution time for GPU contains the memory  transfer.

  Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner products. It compares our CPU library to GMP (where polynomials are  instantiated with VLI or GMP). We get an excellent speed-up factor from 4 to 11, for three reasons. Coefficients of the  polynomial are allocated in the heap, while in the case of GMP, they are allocated in the stack. It is well know that better performances are obtained with less dynamic memory allocation. The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations. Thirdly, the operations are performed in the registers only, no stack call.

The GPU curves show again an excellent speed up  factor of 10 to 60 maximum, compared to GMP, but with an interesting feature. 
For both kinds of polynomials, the performance for one variable is low, because we have a small number of threads during the execution of the GPU algorithms.
As the number of threads is fixed by the number of coefficients, $2n$ for a dense polynomial,  the maximum number of threads will be 28
 which corresponds to one warp. We do not utilize the full capacity of the GPU. For two and more variables, whatever the polynomials, 
  we attain the maximum performance,  if the following conditions are satisfied: The number of running threads is maximum (256 threads), and both input
  polynomials  are cached into the texture memory. On Kepler the size of the texture memory is equal to 48 kB. This memory  is highly optimized for repetitive and random access in contrast to shared memory
 where bank memory conflict can appear.  Moreover, for large order, the resulting polynomials have  more coefficients, so the balance of the tasks over the warp is good, and consequently the efficiency too.

@@

For the small orders the number of GPU threads is lower than 256 but it will increase with the order until the maximum.
After that, for the large orders, although the number of threads is 256, the polynomials will be not full cache into the texture memory. So the execution kernel calls data from the global memory which reduces the performance.
 All this behavior shows the differences between a CPU and a GPU. A CPU has  more transistors for the  data caching and flow control, it ensures a regular and optimal access to the data. For the GPU,
 the programmer manages the flow control by hand. In return, it will have more transistors for the data processing, and consequently more performance.

@@

The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
to GMP for a specific range. The limited performances of GMP for small size were one of the argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate. 
As future work, we would like to implement, new kinds of polynomials, and a better management 
of  the memory to maintain the performance on GPU.

>>>>

%\date{}                                           % Activate to display a given date or no dater
\begin{document}
\maketitle
%\section{}
%\subsection{}
%\vspace{-0.65cm}
\begin{abstract}
The main purpose of the  Very Large Integer (VLI/C++) library is to perform large integer arithmetic and  inner products of polynomials. Based on meta-programming of assembly language,
and  hybrid (CPU/GPU) mode, it performs inner products of dense and triangular polynomials  of 1 to 4 variables  up to order 14 where the coefficients are represented  by large integers of 128 to 512 bits. 
In this paper, we present the basics of the inner products of polynomials, and  the implementation of the library for the CPU and the GPU.  All our algorithms are based on brute force attacks, because the problem is too small for a FFT approach. We save a factor up to 10 on CPU, and 60 on GPU compared to a GMP solution,
and attain 67 \% of the peak performance of the GPU.
\end{abstract}
\section{Introduction}
% TODO references LINBOX, NTL
 A few libraries like  LINBOX or NTL  provide high performance for polynomial solvers.
These libraries utilize with GMP\footnote{GMP is the standard library for long arithmetic, gmplib.org} when needed. 
The introduction of hardware accelerators in HPC a few years ago, and more recently of GPU  by NVIDIA, introduce a revolution in the programming models. 
It is therefore necessary  to adapt the libraries to these new hardwares.  A prominent example of such an adaption are the MAGMA libraries, which port the well known BLAS/LAPACK packages to GPU accelerators.
Polynomials libraries, however,  have not done the transition to GPU yet.
During the last few years, the polynomial multiplication was successfully implemented on this new hardware
\cite{Govindaraju2008, Emeliyanenko:2009, Moreno2010}. All these works focus on large polynomials,
 using a FFT approach.

In the present work, we focus on the inner products of dense and triangular multivariable polynomials.  The coefficients of the polynomials are large integers. 
In contrast to GMP which dynamically adapts the size of the integers, the presented long integer library is based on static integer sizes.
Moreover, our library is natively compatible with GPU to allow hybrid calculation. GPUs impose contiguous memory which is impossible with a GMP approach due to the dynamic allocated memory.
Thus, we propose a  library based on contiguous memory where the exact assembly kernels are generated during the compilation for CPU and GPU.
The static integer sizes allow the compiler to aggressivly optimize the non-assembly parts of the library and avoids costly heap allocations for the data.

@@

The library performs the inner products of vectors of polynomials from one to four variables, where polynomials are defined like (four variables) :
\begin{eqnarray}
P_a(x,y,z,w) & = & \sum_{i=0}^n \sum_{j=0}^n  \sum_{k=0}^n \sum_{l=0}^n  a_{ijkl} x^i y^j z^k w^l , \,\,\,\,\,\, \text{or} \\
                   & = & \sum_{i=0}^n \sum_{j=0}^{(n-i)} \sum_{k=0}^{(n-i-j)} \sum_{l=0}^{(n-i-j-k)}  a_{ijkl} x^i y^j z^k w^l  .
\end{eqnarray}
where  $x,\,y,\,z$ and $w$ are symbolic variables,  $i,\,j,\,k$ and $l$  the indices, $a$  the coefficients, and $n$ the order of the polynomial.
The polynomials of the form  (1) and (2) are named dense and triangular respectively. Consider now two vectors of polynomials of the  same type with $m$ 
entries  $\boldsymbol{v}_1<P_a>$ and  $\boldsymbol{v}_2<P_b>$. The inner products  consist of the product 
of all entries  associated with a global reduction. 
 \begin{eqnarray}
 P_c = \sum_{i=0}^m \boldsymbol{v}_{1_i}<P_a>  \times  \boldsymbol{v}_{2_i}<P_b> \label{poly1} \label{InnerProduct}
\end{eqnarray}
where $\times$ indicates the multiplication between two polynomials. The multiplications use usual rules of polynomials multiplication. 
A final coefficient may be the sum of several coefficients (which we call contributions in the following passage), resulting from the sum of several multiplications. These coefficients are the cross terms. 
The cross terms are not a problem for the CPU version, as a polynomial multiplication is done by a single thread, but they are the bottleneck for the GPU version.
The number of coefficient for a dense and triangular polynomial is given by $(n+1)^{\kappa}$ and  $(\kappa+n-1)! / ( (\kappa-1)! n! )$ respectively, where $\kappa$ is the number of variables.

@@

The coefficients of the polynomials are represented by a large integer named the VLI number.  A VLI number is characterized by a size in bits.
Our current implementation allows for integer sizes from 128 to 512 bits, but is easily extendable to larger integers.
Inside the memory, the VLI numbers are  contiguous. The data type container is a 64 bit unsigned integer array.

The basic operations supported by VLI are the four basic operations: $+,-,\times, /$, extended multiplications and bit shift. The library allows  operations between operands of the same type, and with basic  integers. For performance reasons, we have a specific solver for every operation except  for division.
All the algorithms and operations are based on basic arithmetic but have been implemented in assembly language to make use of hardware features which are not accessible by plain C++.

@@

We applied algorithms based on childhood  experience for large arithmetic. The implementation necessitates a carry bit and an extended multiplication. The hardware supports
the carry bit in the status register, as the extended multiplications, but they are only accessible by the assembly language.

In assembly, these operations  are supported by the pair of mnemonics; \texttt{addq/adcq} for the addition; \texttt{subq/sbbq} for the subtraction, and \texttt{mul} for the extended multiplication.   
Combining these mnemonics, we reproduce easily long arithmetic algorithm, as described into the famous  ASM book of \cite{Hyde:2003:AAL:861534}.

We coded the kernels using the GNU inline assembly.
Generating by hand all these  assembly strings (kernels)  can be painful and error-prone\footnote{The library has near 10000 lines of assembly in total for x86-64, Power64, and PTX}.
We designed a assembly generator. It creates all versions of a kernel during the compilation.   
It is built on Boost preprocessor package. We can thus create and manipulate the ASM strings during the precompilation, 
and generate a specifically tuned kernel for every large integer. We give a small example of a long addition with the generator into the table \ref{ASMGENERATOR}.

@@

The x86-64 architecture allows 16 registers (32 for Power architecture). We try to reuse the data in the registers as much as possible in order to reduce memory read/writes to the minimum.  Additional optimizations have been done like full unrolling and assembly tips. We  supported  x86-64, Power64 and PTX (GPU) assembly language.
 PTX does not support carry bit for 64-bit integers, but only for 32-bit integers\footnote{Introduced since CUDA 4.2}.  
It allows a fused multiplication-addition of 32-bit integer  with carry bit support \cite{CUDAasm}. 
The calculation in 32-bit mode imposes twice as much operations for the additions/subtractions, and one order of magnitude more for the multiplications.

@@

% TODO: which problem are you talking about?
There exists at least two data layouts for this problem: the array of structure (AoS) and structure of array (SoA). Both have  advantages and disadvantages. In our problem the AoS consists in saving the data in this hierarchy: VLI - series of VLI (polynomial coefficient) - series of polynomials (the vector). The SoA structure interleaves the coefficients of the VLI and the polynomials.

@@

The choice of the memory layout is important. It may facilitate  the usage of the SIMD instructions such as the AVX instructions of the Sandy Bridge x86-64 architecture. We choose the AoS for the following reasons:
Applying SIMD to the AoS, for an addition is a nonsense because there is no specific instruction to propagate carry bit between entries of the SIMD register.
The extended multiplication under AVX \texttt{\_\_mm256\_mul\_epu32}  allows only four multiplications of 32-bit, saving a factor of two, nevertheless, on Sandy Bridge, we have 3 ALU, we can not exclude
the execution of several long multiplication at the same time, and we lose the benefit of the SIMD version. Moreover, there is a large overhead  to prepare the registers and shuffle the data into the registers. For our range of operations the extended multiplication varies from 128 to 512 bits, a SIMD version will not be the optimal solution. However SIMD has been used with success for  integers larger  than 1024-bit \cite{SIMD}.
% TODO somehow the problem with SIMD operations is not entirely clear yet. You mention the issues, but I think you should elaborate them a little more.

@@

A Graphical Processing Unit  is efficient  to address problems which can be expressed as data-parallel computations, where the same stream of instructions
is executed on many data elements in parallel. Successful usage of GPU in HPC requires a good understanding of their design to reach great performance. A GPU is composed 
of several independent calculation  entities named streaming multiprocessors (SMs typically 4 to 12), where the SMs run hundreds of threads concurrently (typically 256) in groups of 32 threads (called ``warps'').
The GPU programming model divides the problem into coarse sub-problem that can be solved independently in parallel by blocks of threads,
 and each sub-problem into finer pieces that can be solved cooperatively in parallel by all threads within the block.
% During the execution, independent blocks will be executed on free SMs, where the program of the block is executed over the threads concurrently way.

@@

Evaluating  integer benchmarks  is not easy: unlike the floating point, integer operations are performed by ALU (and not on the single FPU),  the number of  ALUs  between processors (Intel, AMD, IBM) and the execution time of the operations are  not constant \cite{ASMcost} (constant for FPU).  We compute the  result  numbers as operations \footnote{An operation can be a long multiplication, addition and so on}   per seconds $GOP/s=10^{-9} \times batch/t$, where  $batch$ is the number of long integer operations computed during the execution (additions and multiplications), $t$ [s]  is the elapsed time. For more realistic comparison the execution time for GPU contains the memory  transfer.

Figure \ref{ResME} presents the performance of a dense and triangular polynomials inner products. It compares our CPU library to GMP (where polynomials are  instantiated with VLI or GMP). We get an excellent speed-up factor from 4 to 11, for three reasons.
Coefficients of the  polynomial are allocated in one large block on the heap, while in the case of GMP, each coefficient needs to be allocated independently. It is well known that better performances are obtained with less dynamic memory allocations. The second reason is that  the meta-programming allows the construction of a specific solver for every type of large integer during the compilation, thus, we do not have a universal solver, the solver being unique with the maximum optimizations. Thirdly, the operations are performed in the registers only, no stack call.
%TODO what do you mean by the last sentence?

The GPU curves show again an excellent speed up  factor of 10 to 60 maximum, compared to GMP, but with an interesting feature. 
For both kinds of polynomials, the performance for one variable is low, because we have a small number of threads during the execution of the GPU algorithms.
As the number of threads is fixed by the number of coefficients, $2n$ for a dense polynomial,  the maximum number of threads will be 28
 which corresponds to one warp. We do not utilize the full capacity of the GPU. For two and more variables, whatever the polynomials, 
  we attain the maximum performance,  if the following conditions are satisfied: The number of running threads is maximum (256 threads), and both input
  polynomials  are cached into the texture memory. On Kepler the size of the texture memory is equal to 48 kB. This memory  is highly optimized for repetitive and random access in contrast to shared memory
 where bank memory conflicts may appear.
Moreover, for high orders the resulting polynomials have more coefficients which eases the the work balance between different threads and consequently improves efficiency.

@@

For the small orders the number of GPU threads is lower than 256 but it will increase with the order until the maximum.
After that, for the large orders the polynomials will be too large to be cached by texture memory properly. So the execution kernel calls data from the global memory which reduces the performance.
 All this behavior shows the differences between a CPU and a GPU. A CPU has  more transistors for the  data caching and flow control, it ensures a regular and optimal access to the data. For the GPU,
 the programmer manages the flow control by hand. In return, it will have more transistors for the data processing, and consequently more performance.

@@

The VLI library provides efficient solvers for the 128 to 512 bits integer associated with polynomials of 1 to 4 variables. We obtain a comfortable speed up as high as  60 compared 
to GMP for a specific range. The limited performances of GMP for small size were the main argument to develop the VLI library, although GMP excels in thousand bits arithmetic where our library can not operate. 
As future work, we would like to implement, new kinds of polynomials, and a better management 
of  the memory to maintain the performance on GPU.
